---
output: 
  github_document:
    toc: true
    toc_depth: 3
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# HOQCenc

<!-- badges: start -->
<!-- badges: end -->

HOQCenc contains routines for encryption and decryption of character strings and files. The main function is `xcode` that encrypts an input string by executing a number of operations (specified by argument `trans`) when argument `ed='e'`. When `ed='d'` is specified decryption will be done by executing the operations in the reversed order.

## Installation

You can install this version from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("HanOostdijk/HOQCenc") 
```

```{r echo=F}
library(HOQCenc)
```

## Example
In the [Details](#details) section more examples are given and some background. In this first example we show:

- how to encrypt a message
- that the result of encryption always is a character string with letters from the [alphabet](#key-and-alphabet)
- that by choosing a set of [Operations](#operations) it is possible that small changes in the message lead to big differences in the encrypted text 
- that (happily) a sequence of encrypting and decrypting returns the original string

```{r}
key <- "1VerySecretKey"
message1 <- "Line 1 of a two-line message!\nAs expected a second line."
message2 <- "Line 1 of a two-line message.\nAs expected a second line." 

(s <- xcode(message1,key=key,ed='e',trans='cscvp')) # encrypt message1
xcode(s,key=key,ed='d',trans='cscvp')               # show result of encryption
xcode(message2,key=key,ed='e',trans='cscvp')        # encrypt message2
```
 
## Details {#details}

### Key and alphabet
When we specify `key=''` an ordered set is constructed consisting of the lower and upper case letters, the digits 0 till 9 and the space and the `#` character. This ordered set is our 'alphabet': all operations will work only with this set with the exception of the `e` operation that converts our standard symbols to this new alphabet.  
When we use the key `1VerySecretKey` the characters in this key will be inserted in this set before the others so that the following ordered set will be constructed:

```r
     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8]
[1,] "1"  "V"  "e"  "r"  "y"  "S"  "c"  "t" 
[2,] "K"  "a"  "b"  "d"  "f"  "g"  "h"  "i" 
[3,] "j"  "k"  "l"  "m"  "n"  "o"  "p"  "q" 
[4,] "s"  "u"  "v"  "w"  "x"  "z"  "A"  "B" 
[5,] "C"  "D"  "E"  "F"  "G"  "H"  "I"  "J" 
[6,] "L"  "M"  "N"  "O"  "P"  "Q"  "R"  "T" 
[7,] "U"  "W"  "X"  "Y"  "Z"  "0"  "2"  "3" 
[8,] "4"  "5"  "6"  "7"  "8"  "9"  " "  "#" 
```
In almost all operations (but not the `e`, `s` and `f` operation) this 'alfabet' will be used.

### Operations {#operations}
The following operations are defined (we describe only what happens in case of encryption
 because for decryption the reversed action is performed):

- e : encode all characters that are not a blank, a letter or a digit
 and ensure that the length of the resulting string is a multiple of 16
- s : shuffle the characters in such a way that `0123456789abcdef` is translated to `0f2d4b6987a5c3e1`
- f : flip the characters in such a way that `0123456789abcdef` is translated to `fedcba9876543210`
- v : a Vigenère translation based on the key: the x-th character is shifted in the ordered set
based on the position of the x-th character in the ordered set.
- c : a Vigenère translation based on the first character of the ordered set for the first character
and from then on based on the result for the preceding character.
- p : a Playfair translation based on an 8x8 square filled with the ordered set (as show above)
- a : an AES translation based on the key. This is in fact the [digest::AES()] function with `mode='ECB'`

The encryption characteristics of these operations are the following:

- e : no real encryption because with a hex table the source text is easily derived
- s : no encryption but useful in combination
- f : no encryption but useful in combination
- v : encryption with strength 'proportional' to the number of unique characters in the key
- c : encryption based on key of length one, but has the advantage that a difference in a character leads
to difference in later results. Always combine with `v`, `p` or `a` operations
- p : pairwise encryption with strength better when the number of unique characters in the key is greater.
When a pair occurs more than once, it is encrypted the same each time
- a : When a 16-character block occurs more than once, it is encrypted the same each time.

So it is advised to combine several options as e.g. done in the example below



```{r}
key <- "1VerySecretKey"

# small differences in text lead to similar outcomes with simple trans
xcode("abcdefghijklmnopqrstuwvxyz",key=key,ed='e',trans='p')
xcode("1bcdefghijklmnopqrstuwvxyz",key=key,ed='e',trans='p')
xcode("abcdefghijklmnopqrstuwvxy1",key=key,ed='e',trans='p')
xcode("ABabcdefghijklmnopqrstuwvxy1",key=key,ed='e',trans='p')

# small differences in text lead to unrelated outcomes with complex trans
xcode("abcdefghijklmnopqrstuwvxyz",key=key,ed='e',trans='cscvp')
xcode("1bcdefghijklmnopqrstuwvxyz",key=key,ed='e',trans='cscvp')
xcode("abcdefghijklmnopqrstuwvxy1",key=key,ed='e',trans='cscvp')
xcode("ABabcdefghijklmnopqrstuwvxy1",key=key,ed='e',trans='cscvp')

```


### The `e` operation
Most of the operations used in the `xcode` function are working only at the 64 characters of our ordered set of the lower and upper case letters, the digits 0 till 9 and the space and the `#` character. 
The `e` operation: 

- encodes the `#` character and the characters outside the ordered set by replacing each of those by `#` and their hex representation. E.g. the exclamation mark is replaced by `#21` and the `#` character itself is replaced by `#23` .
- ensures that the number of characters is a multiple of 16 by adding one or more fillers (consisting of `#00`) and possible one or two spaces. The multiple of 16 is necessary when the `a` operation (AES encryption) is used but for the other operations an even number of characters is sufficient.

Because the `e` operation is necessary for most cases it is always prefixed to the list of operations that is specified in the `trans` argument. 

```{r error=T}
xcode('Deze #!',key='',ed='e',trans='')
```
```{r}
xcode('example!',key='',ed='e',trans='')
xcode('example!',key='',ed='e',trans='p')
xcode('example!',key='',ed='e',trans='a')
```
Only when the `noe` argument is explicitly set to `TRUE` this is not done.
```{r error=T}
xcode('example!',key='',ed='e',trans='',noe=T)
xcode('example!',key='',ed='e',trans='p',noe=T)
xcode('example ',key='',ed='e',trans='p',noe=T) # but with space instead of !
xcode('example ',key='',ed='e',trans='a',noe=T)
```

### The `p` operation 
With the `p` operation a Playfair transformation is done that acts on pairs of characters.
With `key=''` the following 'alfabet' is used
```r
     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8]
[1,] "a"  "b"  "c"  "d"  "e"  "f"  "g"  "h" 
[2,] "i"  "j"  "k"  "l"  "m"  "n"  "o"  "p" 
[3,] "q"  "r"  "s"  "t"  "u"  "v"  "w"  "x" 
[4,] "y"  "z"  "A"  "B"  "C"  "D"  "E"  "F" 
[5,] "G"  "H"  "I"  "J"  "K"  "L"  "M"  "N" 
[6,] "O"  "P"  "Q"  "R"  "S"  "T"  "U"  "V" 
[7,] "W"  "X"  "Y"  "Z"  "0"  "1"  "2"  "3" 
[8,] "4"  "5"  "6"  "7"  "8"  "9"  " "  "#" 
```
and therefore "ex" is converted to "hu' because 'h' and 'u' lie on the opposite corner of the rectangle though 'e' and 'x'. Special cases:

- equal row : 'km' -> 'ln'  and  'ip' -> 'ji
- equal column : 'zP' -> 'HX' and 'J7' -> 'Rd'
- equal cell : 'CC' -> 'LL' , 'FF' -> 'GG' and '##' -> 'aa'

Advantage of Playfair is that a frequency distribution on character level makes little sense because the method works on pairs of characters. 
```{r}
xcode('example ',key='',ed='e',trans='p',noe=T)
xcode("hueiimg8",key='',ed='d',trans='p',noe=T)
```

### The `s` and `f` operations
The `s` (shuffle) operation shuffles the characters in such a way that `0123456789abcdef` is translated to `0f2d4b6987a5c3e1` and the `f` operations (flip) flips the characters in such a way that `0123456789abcdef` is translated to `fedcba9876543210`. 

```{r}
xcode('0123456789abcdef',key='',ed='e',trans='s',noe=T)
xcode('0123456789abcdef',key='',ed='e',trans='f',noe=T)
```

### The `v` operation
The `v` (Vigenère) operation shifts the i-th character a number of places in the 'alfabet'. That number is determined by the position of the i-th letter in the 'alfabet'. In the first example the 'alfabet' is `ABCabcdef ... ` and therefore the shifts for the letters 'a', 'b' etc. are 4, 5 etc. resulting in '4', '6' etc. In the second example the 'alfabet' is `abcdef ... ` and therefore the shifts for the letters 'a', 'b' etc. are 1, 2 etc. resulting in '1', '3' etc.

```{r}
xcode('0123456789abcdef',key='ABC',ed='e',trans='v',noe=T)
xcode('0123456789abcdef',key='abc',ed='e',trans='v',noe=T)
```

### The `c` operation
The `c` operation shifts the first character a number of places in the 'alfabet'. That number is again determined by the position of the first letter in the 'alfabet'. In the first example the 'alfabet' is `abcdef ... ` and therefore the shifts for the letter 'a' is 1 resulting in '1' . The shift for the next position is therefore determined by the position of '1' in the 'alfabet' and that is 44. So the next position (the '1') is shifted by 44 positions in the 'alfabet' resulting in 'R' . And so on ...
The main advantage of this method is that the encryption of a character is determined by all of its predecessors. The weak point is that only the position of the 'a' in the 'alfabet' has to be guessed or tried by brute force and this is very easy. So this method should never be used on its own.

```{r}
xcode('0123456789abcdef',key='abc',ed='e',trans='c',noe=T)
```

### The `a` operation
The `a` uses the `digest::AES` function with the `mode='ECB'` method. The method works in blocks of 16 characters: blocks that are identical with the exception of one character will be encrypted very unrelated results. However if a block occurs more than once, each block is identically encryped. So also this method should never be used on its own.

```{r}
xcode('0123456789abcdef',key='abc',ed='e',trans='a',noe=T)
xcode('1123456789abcdef',key='abc',ed='e',trans='a',noe=T)
xcode('0123456789abcdee',key='abc',ed='e',trans='a',noe=T)

xcode('0123456789abcdef0123456789abcdef',key='abc',ed='e',trans='a',noe=T)
```

